Determinism, Complexity, and Predictability in Computer Performance

### 2014-04-02 intro.tex
edits and consistency changes

In particular, we wish to determine---without knowing anything about
the generating process---whether a time series is too complex to
predict.  An important practical corollary to this is a strategy for
assessing appropriateness of forecast methods.  If the forecast
produced by a particular method is poor but the time series contains a
significant amount of predictive structure, one can reasonably conclude that that method is inadequate
to the task and that one should seek another method.

@@

We focus on real-valued, scalar, time-series data that appear
complicated: i.e., not periodic, linear, constant, etc.  We make no
assumptions about the properties of the generating process: whether it
is linear, nonlinear, deterministic, stochastic, stationary,
non-stationary, etc.
To explore the relationship between complexity, predictive structure,
and predictability, we generate forecasts of a variety of time-series
datasets using four different prediction methods, then compare the
accuracy of those predictions to the permutation entropy of the
associated signals.  This results in two primary findings:
\begin{enumerate}
\item The complexity of a noisy real-valued time series is {\sl (i)}
  quantifiable by permutation entropy and {\sl (ii)} correlated with prediction
  accuracy of an appropriate predictor.
\item The way information is generated and processed internally by a
  system plays a crucial role in the success of different forecasting
  schema---and in the choice of which one is appropriate for a given 
time series.
\end{enumerate}
The forecast methods used in this study were chosen to be a
representative sampling of standard prediction strategies, but they do
not, of course, cover that space exhaustively.  Our goal here is an
empirical assessment of the relationship between predictability and
complexity, not formal results about a ``best'' predictor for a given
time series.  From a practitioner's standpoint, it would be useful to
know {\sl a priori} if a trace contains enough predictive structure to
make it worth spending the time and effort searching for a good
forecast method.  It would also be useful to know if a given method is
inadequate---that is, if one could do better and should keep
searching.

@@

For the purposes of this study, we require a broad array of
time-series datasets from across the complexity spectrum.  Rather than
use synthetic data, we study sensor data from computer performance
experiments.  While this is not a common laboratory experiment, it is
a highly appropriate choice here.  Computers are among the most
complex engineered artifacts in current use.  Modern microprocessor
chips contain multiple processing units and multi-layer memories, for
instance, and they use complicated hardware/software strategies to
move data and threads of computation across those resources.  As a
result, the processor and memory loads during the execution of even a
very simple program can evolve in a very complicated fashion.
Figure~\ref{fig:col-ipc} shows a performance trace of a three-line
loop that repeatedly initializes the upper triangle of a matrix in
column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a three-line C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:col-ipc}
  \end{figure}
%
Although this time series appears to be periodic, it is actually
chaotic~\cite{mytkowicz09}, which places fundamental limits on
predictability.  Changes in the code can radically change this
behavior; indeed, many performance traces exhibit interesting regime
changes as a program moves through the different phases of its
operation.  The resulting traces span the whole range of the
complexity spectrum, from completely predictable to completely
unstructured, making this an ideal testbed for this
study\footnote{Predicting the \emph{state} of a computer, of course,
  would amount to solving the halting problem.  What we are doing here
  is predicting computer \emph{performance}, which does not violate
  the Rice-Shapiro theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

>>>>

In particular, we wish to determine---without knowing anything about
the generating process---whether a time series is too complex to successfully
predict.  An important practical corollary to this is a strategy for
assessing appropriateness of forecast methods.  If the forecast
produced by a particular method is poor but the time series contains a
significant amount of predictive structure, one can reasonably conclude that that method is inadequate
to the task and that one should seek another method.

@@

We focus on real-valued, scalar, time-series data that appear
complicated: i.e., not periodic, linear, constant, etc.  We make no
assumptions about the properties of the generating process: whether it
is linear, nonlinear, deterministic, stochastic, stationary,
non-stationary, etc.
To explore the relationship between complexity, predictive structure,
and predictability, we generate forecasts of a variety of time-series
datasets using four different prediction methods, then compare the
accuracy of those predictions to the permutation entropy of the
associated signals.  This results in two primary findings:
\begin{enumerate}
\item The complexity of a noisy real-valued time series is {\sl (i)}
  quantifiable by permutation entropy and {\sl (ii)} correlated with prediction
  accuracy of an appropriate predictor.
\item The way information is generated and processed internally by a
  system plays a crucial role in the success of different forecasting
  schema---and in the choice of which one is appropriate for a given 
time series.
\end{enumerate}
The forecast methods used in this study were chosen to be a
representative sampling of standard prediction strategies, but they do
not, of course, cover that space exhaustively.  Our goal here is an
empirical assessment of the relationship between predictability and
complexity, not formal results about a ``best'' predictor for a given
time series.  From a practitioner's standpoint, it would be useful to
know {\sl a priori} if a time series contains enough predictive structure to
make it worth spending the time and effort searching for a good
forecast method.  It would also be useful to know if a given method is
inadequate---that is, if one could do better and should keep
searching.

@@

For the purposes of this study, we require a broad array of
time-series datasets from across the complexity spectrum.  Rather than
use synthetic data, we study sensor data from computer performance
experiments.  While this is not a common laboratory experiment, it is
a highly appropriate choice here. {\color{red} [[Liz, these are the few sentences I was worried about being too similar but we may have only used them in a grant proposal.]]} Computers are among the most
complex engineered artifacts in current use.  Modern microprocessor
chips contain multiple processing units and multi-layer memories, for
instance, and they use complicated hardware/software strategies to
move data and threads of computation across those resources.  As a
result, the processor and memory loads during the execution of even a
very simple program can evolve in a very complicated fashion.
Figure~\ref{fig:col-ipc} shows a performance trace of a three-line
loop that repeatedly initializes the upper triangle of a matrix in
column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a three-line C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:col-ipc}
  \end{figure}
%
Although this time series appears to be periodic, it is actually
chaotic~\cite{mytkowicz09}, which places fundamental limits on
predictability.  Changes in the code can radically change this
behavior; indeed, many performance traces exhibit interesting regime
changes as a program moves through the different phases of its
operation.  The resulting traces span the whole range of the
complexity spectrum, from completely predictable to completely
unstructured, making this an ideal testbed for this
study\footnote{Predicting the \emph{state} of a computer, of course,
  would amount to solving the halting problem.  What we are doing here
  is predicting computer \emph{performance}, which does not violate
  the Rice-Shapiro theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

### 2014-04-02 methods.tex
edits and consistency changes

The time-series data for these experiments was collected on an Intel
Core\textsuperscript{\textregistered} i7-2600-based machine running
the 2.6.38-8 Linux kernel.  This particular microprocessor chip has
eight processing units, a clock rate of 3.40GHz, and a cache size of
8192 KB.  We gathered performance traces during the execution of three
different programs---the simple \col loop whose performance is
depicted in Figure~\ref{fig:col-ipc}, as well as two more-complex
programs: one from the SPEC 2006CPU benchmark suite (\gcc), and one
from the LAPACK linear algebra package (\svd).  In these experiments,
the scalar observation $x_{i}$ at time $i$ was a measurement of how
well each program was able to exploit the native parellelism of the i7
chip.  To record these measurements, we used the {\tt libpfm4}
library, via PAPI
% 
% \footnote{Performance Application Programming Interface}
% 
5.2~\cite{papi}, to stop program execution at 100,000-instruction
intervals and read the contents of the CPU's onboard hardware
performance monitors\footnote{These specialty registers are built into
  modern microprocessors for the purpose of on-board monitoring and
  storage of performance data.  They can be programmed to measure a
  wide array of processor and memory metrics besides the one used
  here.}, which we had programmed to count how many instructions were
executed in each clock cycle (IPC).  For an in-depth description of
this custom-measurement infrastructure, including discussion of the
implications of the sampling interval, please see
\cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical validation,
we collected 15 performance traces from each of the three programs.
These traces, and the processes that generated them, are described in
more depth in the rest of this section.

@@

\col is a simple three-line C program that repeatedly initializes the
upper triangle of a 2048 $\times$ 2048 matrix in column-major order.
As mentioned in the previous section, this simple program exhibits
surprisingly complicated behavior~\cite{mytkowicz09}.  A time series
of the processor utilization during the execution this program is
shown in Figure~\ref{fig:col-ts}.
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/colFullTS}
    \caption{The instructions executed per CPU clock cycle (IPC)
      during the execution of \col.  Each point is the average IPC
      during a 100,000 instruction period. Figure~\ref{fig:col-ipc} is
      a closeup of the first part of this signal.  }
    \label{fig:col-ts}

@@

\svd is a Fortran program from the LAPACK linear algebra package
\cite{lapack}.  It calculates the singular value decomposition of an
rectangular $M$ by $N$ real-valued matrix.  For our experiments, we
choose $M=750$ and $N=1000$ and generated the matrix entries randomly.
% 
% \footnote{Multiple randomly generated matrices were
%   investigated but no measurable effect was present in the resulting
%   time series.}. 
% 
The behavior of this program as it computes the singular values of
this matrix is very interesting, as shown in Figure
\ref{fig:svd-ts-colored}.  As the code moves though its different
phases---diagonalizing the matrix, computing its transpose,
multiplying, etc.---the processor utilization changes quite radically,
as shown in Figure~\ref{fig:svd-ts-colored}.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/SVD1RegimesColored}
    \caption{The instructions executed per CPU clock cycle (IPC)
      during the execution of \svd. Each point is the average IPC
      during a 100,000 instruction period.  The colors refer to
      different {\sl segments} of the signal, as discussed in the
      text.}
    \label{fig:svd-ts-colored}
  \end{figure}
For the first $\sim$21,000 measurements (21,000 $\times$ 100,000
cycles), roughly 1.8 instructions are executed per cycle, on the
average, by the eight processing units on this chip.  After that, the
IPC moves through a number of different oscillatory regimes, which we
have color-coded in the figure in order to make textual
cross-references easy to track and understand.

This wide range of behaviors provides a unique advantage, for the
purposes of this paper, in that a number of different generating
processes---with a wide range of complexities---are at work at
different points in a single time series.  The complexity of \col and
\gcc in Figures~\ref{fig:col-ts} and \ref{fig:gcc-ts} appears to be
far more consistent over time---probably the result of a single
generating process.  \svd, in contrast, has multiple regimes, each the
result of a different generating process.  To take advantage of this,
we split the signal into six different segments, thereby obtaining an
array of examples for the analyses in the following sections.  For
notational convenience, we refer to these 90 time series\footnote{15
  runs, each with six regimes} as {\tt dgesdd$_i$} with $i \in
\{1\dots6\}$ where $i$ corresponds to one of the six color-coded
segments of \svd, ordered from left to right.  These segments, which
were determined visually, are shown in different colors in
Figure~\ref{fig:svd-ts-colored}.  Visual decomposition is subjective,
of course, particularly since the regimes exhibit some fractal
structure.  Thus, it may the case that more than one generating
process may be at work in each of our segments.  This is a factor in
the discussion of Section~\ref{sec:results}.

>>>>

The time-series data for these experiments was collected on an Intel
Core\textsuperscript{\textregistered} i7-2600-based machine running
the 2.6.38-8 Linux kernel.  This particular microprocessor chip has
eight processing units, a clock rate of 3.40GHz, and a cache size of
8192 KB.  We gathered performance traces during the execution of three
different programs---the simple \col loop whose performance is
depicted in Figure~\ref{fig:col-ipc}, as well as two more-complex
programs: one from the SPEC 2006CPU benchmark suite (\gcc), and one
from the LAPACK linear algebra package (\svd).  In these experiments, the scalar observation $x_i$ was a measurement of the processor performance at time $i$ during the execution of each program.  To record these measurements, we used the {\tt libpfm4}
library, via PAPI
% 
% \footnote{Performance Application Programming Interface}
% 
5.2~\cite{papi}, to stop program execution at 100,000-instruction
intervals and read the contents of the CPU's onboard hardware
performance monitors\footnote{These specialty registers are built into
  modern microprocessors for the purpose of on-board monitoring and
  storage of performance data.  They can be programmed to measure a
  wide array of processor and memory metrics besides the one used
  here.}, which we had programmed to count how many instructions were
executed in each clock cycle (IPC).  For an in-depth description of
this custom-measurement infrastructure, including discussion of the
implications of the sampling interval, please see
\cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical validation,
we collected 15 performance traces from each of the three programs.
These traces, and the processes that generated them, are described in
more depth in the rest of this section.

@@

\col is a simple three-line C program that repeatedly initializes the
upper triangle of a 2048 $\times$ 2048 matrix in column-major order.
As mentioned in the previous section, this simple program exhibits
surprisingly complicated behavior~\cite{mytkowicz09}.  A time series
of the processor performance during the execution this program is
shown in Figure~\ref{fig:col-ts}.
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/colFullTS}
    \caption{The instructions executed per CPU clock cycle (IPC)
      during the execution of \col.  Each point is the average IPC
      during a 100,000 instruction period. Figure~\ref{fig:col-ipc} is
      a closeup of the first part of this signal.  }
    \label{fig:col-ts}

@@

\svd is a Fortran program from the LAPACK linear algebra package
\cite{lapack}.  It calculates the singular value decomposition of a
rectangular $M$ by $N$ real-valued matrix.  For our experiments, we
choose $M=750$ and $N=1000$ and generated the matrix entries randomly.
% 
% \footnote{Multiple randomly generated matrices were
%   investigated but no measurable effect was present in the resulting
%   time series.}. 
% 
The behavior of this program as it computes the singular values of
this matrix is very interesting, as shown in Figure
\ref{fig:svd-ts-colored}.  As the code moves though its different
phases---diagonalizing the matrix, computing its transpose,
multiplying, etc.---the processor utilization changes quite radically,
as shown in Figure~\ref{fig:svd-ts-colored}.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/SVD1RegimesColored}
    \caption{The instructions executed per CPU clock cycle (IPC)
      during the execution of \svd. Each point is the average IPC
      during a 100,000 instruction period.  The colors refer to
      different {\sl segments} of the signal, as discussed in the
      text.}
    \label{fig:svd-ts-colored}
  \end{figure}
For the first $\sim$21,000 measurements (21,000 $\times$ 100,000
instructions), roughly 1.8 instructions are executed per cycle, on the
average, by the eight processing units on this chip.  After that, the
IPC moves through a number of different oscillatory regimes, which we
have color-coded in the figure in order to make textual
cross-references easy to track and understand.

This wide range of behaviors provides a unique advantage, for the
purposes of this paper, in that a number of different generating
processes---with a wide range of complexities---are at work at
different points in a single time series.  The complexity of \col and
\gcc in Figures~\ref{fig:col-ts} and \ref{fig:gcc-ts} appears to be
far more consistent over time---probably the result of a single
generating process.  \svd, in contrast, has multiple regimes, each the
result of a different generating process.  To take advantage of this,
we split the signal into six different segments, thereby obtaining an
array of examples for the analyses in the following sections.  For
notational convenience, we refer to these 90 time series\footnote{15
  runs, each with six regimes} as {\tt dgesdd$_i$} with $i \in
\{1\dots6\}$ where $i$ corresponds to one of the six color-coded
segments of \svd, ordered from left to right.  These segments, which
were determined visually, are shown in different colors in
Figure~\ref{fig:svd-ts-colored}.  Visual decomposition is subjective,
of course, particularly since the regimes exhibit some fractal
structure.  Thus, it may be the case that more than one generating
process may be at work in each of our segments.  This is a factor in
the discussion of Section~\ref{sec:results}.
